---
title: "COVID-19 Data Analysis Visualizations"
author: "Youssef Khabir"
date: "November 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 6, dpi = 300)

# Load required libraries with conflict resolution
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
library(stringr)

# Set global options for better output
options(scipen = 999, stringsAsFactors = FALSE)
```

# Data Loading and Processing

```{r data-prep, include=FALSE}
# Import the dataset
covid_data <- read_csv("data/raw/owid-covid-data.csv", show_col_types = FALSE)

# Data cleaning and preprocessing (simplified version)
covid_data <- covid_data %>%
  mutate(date = as.Date(date)) %>%
  dplyr::filter(!is.na(location), !is.na(date)) %>%
  dplyr::filter(location != "International" & location != "World") %>%
  mutate(
    total_cases = ifelse(is.na(total_cases), 0, total_cases),
    new_cases = ifelse(is.na(new_cases), 0, new_cases),
    total_deaths = ifelse(is.na(total_deaths), 0, total_deaths),
    new_deaths = ifelse(is.na(new_deaths), 0, new_deaths),
    case_fatality_rate = ifelse(total_cases > 0, (total_deaths / total_cases) * 100, 0),
    vaccination_coverage = ifelse(population > 0, (people_fully_vaccinated / population) * 100, 0)
  )
```

# Visualization 1: Global Daily New COVID-19 Cases

```{r global-cases}
# Global cases over time
global_timeline <- covid_data %>%
  group_by(date) %>%
  summarise(
    new_cases = sum(new_cases, na.rm = TRUE),
    .groups = 'drop'
)

ggplot(global_timeline, aes(x = date)) +
  geom_line(aes(y = new_cases/1000), color = "steelblue", size = 1) +
  geom_smooth(aes(y = new_cases/1000), method = "loess", se = FALSE, color = "red") +
  labs(
    title = "Global Daily New COVID-19 Cases",
    subtitle = "Thousand cases per day with trend line",
    x = "Date",
    y = "New Cases (thousands)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Key Insights:
- Distinct pandemic waves visible over time
- Peak periods correspond to variant emergence and spread
- Trend line shows overall pattern of the pandemic progression

---

# Visualization 2: Top 10 Countries by Total Cases

```{r top-countries}
# Get top countries
top_countries <- covid_data %>%
  group_by(location) %>%
  summarise(
    total_cases = max(total_cases, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_cases)) %>%
  head(10)

ggplot(top_countries, aes(x = reorder(location, total_cases), y = total_cases/1000000)) +
  geom_col(fill = "coral", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Total COVID-19 Cases",
    subtitle = "As of latest data available",
    x = "Country",
    y = "Total Cases (millions)"
  ) +
  theme_minimal()
```

## Key Insights:
- United States, India, and Brazil lead in total case counts
- Population size correlates with total case numbers
- European countries prominently represented in top 10

---

# Visualization 3: Case Fatality Rate by Continent

```{r cfr-continent}
# Case fatality rate by continent
continent_cfr <- covid_data %>%
  dplyr::filter(!is.na(continent), total_cases > 1000) %>%
  group_by(continent) %>%
  summarise(
    avg_cfr = mean(case_fatality_rate, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(continent_cfr, aes(x = reorder(continent, avg_cfr), y = avg_cfr)) +
  geom_col(fill = "forestgreen", alpha = 0.7) +
  labs(
    title = "Average Case Fatality Rate by Continent",
    subtitle = "For countries with 1000+ cases",
    x = "Continent",
    y = "Average CFR (%)"
  ) +
  theme_minimal() +
  coord_flip()
```

## Key Insights:
- Relatively consistent CFR across major continents (2-3%)
- Variations reflect differences in healthcare capacity and population demographics
- Lower rates in regions with younger population structures

---

# Visualization 4: COVID-19 Cases vs Population Density

```{r density-analysis}
# Cases vs Population Density
density_analysis <- covid_data %>%
  dplyr::filter(!is.na(population_density), total_cases_per_million > 0, population_density < 1000) %>%
  group_by(location) %>%
  summarise(
    cases_per_million = max(total_cases_per_million, na.rm = TRUE),
    population_density = max(population_density, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(density_analysis, aes(x = population_density, y = cases_per_million)) +
  geom_point(alpha = 0.4, color = "purple") +
  geom_smooth(method = "lm", se = FALSE, color = "orange", size = 1) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "COVID-19 Cases vs Population Density",
    subtitle = "Log-log scale relationship",
    x = "Population Density (people per km², log scale)",
    y = "Cases per Million (log scale)"
  ) +
  theme_minimal()
```

## Key Insights:
- Log-log scale reveals complex relationship between density and transmission
- Higher density generally associated with higher per-capita cases
- Significant variation suggests other important factors at play

---

# Visualization 5: Global COVID-19 Vaccination Progress

```{r vaccination-progress}
# Vaccination progress over time
vaccination_timeline <- covid_data %>%
  dplyr::filter(date >= "2021-01-01") %>%
  group_by(date) %>%
  summarise(
    avg_vaccination_coverage = mean(vaccination_coverage, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(vaccination_timeline, aes(x = date, y = avg_vaccination_coverage)) +
  geom_line(color = "green", size = 1) +
  geom_area(fill = "green", alpha = 0.3) +
  labs(
    title = "Global COVID-19 Vaccination Progress",
    subtitle = "Average vaccination coverage over time",
    x = "Date",
    y = "Vaccination Coverage (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Key Insights:
- Rapid vaccination uptake from 2021 onwards
- Clear acceleration phase during global vaccine rollout
- Plateau phase as coverage approaches saturation levels

---

# Summary

This R Markdown document generates all five key visualizations for the COVID-19 analysis:

1. **Global Daily Cases**: Shows pandemic waves and trends over time
2. **Top Countries**: Compares total case counts across most affected nations
3. **Case Fatality Rates**: Analyzes mortality outcomes by geographic region
4. **Population Density**: Explores relationship between crowding and transmission
5. **Vaccination Progress**: Tracks global immunization campaign

Each visualization includes professional styling, clear labeling, and explanatory insights to support comprehensive COVID-19 analysis and reporting.